---
title: "CleanData"
author: "Dr. B"
date: "Wednesday, November 26, 2014"
output: html_document
---


```{r, warning=FALSE, message=FALSE}
##Use my standard openning including call function
source('C:/Users/bryan_000/Documents/GitHub/MyWork/StdOpen.R')
```


```{r,results='markup'}
##Load some data 
data <-read.table("http://archive.ics.uci.edu/ml/machine-learning-databases/pima-indians-diabetes/pima-indians-diabetes.data", header=F, sep=",")

##create columnames that need to be tidy
colnames (data) <- c("N_Preg", "glu//cose","BP", "tri-ceps", "Ins ulin", "bmi", "diabetes", "AGE", "clas s ")

#copy data to df
df <-data

##View summary Check for NA
summary(df)
apply(is.na(df),2,sum)
```

Variable "npreg" has maximum value equal to 17, unusually large but not impossible

Variables "glucose", "bp", "triceps", "insulin" and "bmi" have minimum value equal to zero      in this case, it seems that zero was used to code missing data

Zero should not be used to represent missing data: (1) it's a valid value for some of the variables, (2) can yield misleading results.

Variable "class" is categorical, not quantitative.  Categorical should not be coded as numerical data
        
```{r cleanup}
# Clean data column names for the data frame
names(df)<- tolower(names(df))
names(df)<- gsub("-","",names(df))
names(df)<- gsub("_","",names(df))
names(df)<- gsub(" ","",names(df))
#names(df)<- gsub("\\","",names(df))
names(df)<- gsub("//","",names(df))

###Set the missing values erronusly coded as zero to NA
df$glucose[df$glucose==0] <- NA
df$bp[df$bp==0] <- NA
df$triceps[df$triceps==0] <- NA
df$insulin[df$insulin==0] <- NA
df$bmi[df$bmi==0] <- NA

##Set categorical variables coded as numerical to factor
#df$class <- factor(df$class) 
#levels(df$class) <- c("neg", "pos")

##Summary & check for N/A
summary(df)
apply(is.na(df),2,sum)

# impute with mean value
df$glucoseimputed <- with(df, impute(glucose, mean))
df$bpimputed <- with(df, impute(bp, mean))
df$tricepsimputed <- with(df, impute(triceps, mean))
df$insulinimputed <- with(df, impute(insulin, mean))
df$bmiimputed <- with(df, impute(bmi, mean))

df$glucose<- NULL
df$bp<- NULL
df$triceps <- NULL
df$insulin <- NULL
df$bmi <- NULL

##Convert all imputed variables to nummeric
#df$glucose <-as.numeric(df$glucose)
#df$bp <- as.numeric(df$bp)
#df$triceps <- as.numeric(df$triceps)
#df$insulin <- as.numeric(df$insulin)
#df$bmi <- as.numeric(df$bmi)

##Summary & check for N/A
summary(df)
apply(is.na(df),2,sum)
```


```{r}
splits<-dfsplit(df)
df.train<-splits$trainset
df.test<-splits$testset
```


```{r visual}
##Set printing
par(mfrow=c(1,2))

hist(df$npreg)
qqnorm(df$npreg, main="QQ Plot \n NPreg Variable")
qqline(df$npreg)

hist(df$glucose)
qqnorm(df$glucose, main="QQ Plot \n Glucose Variable")
qqline(df$glucose)

hist(df$bp)
qqnorm(df$bp, main="QQ Plot \n DP Variable")
qqline(df$bp)

hist(df$diabetes)
qqnorm(df$diabetes, main="QQ Plot \n Diabetes Variable")
qqline(df$diabetes)

hist(df$triceps)
qqnorm(df$triceps, main="QQ Plot \n Triceps Variable")
qqline(df$triceps)

hist(df$insulin)
#qqnorm(df$insulin, main="QQ Plot \n Insulin Variable")
#qqline(df$insulin)
```

Logistic Regession
```{r}
##Regression model
basefit <-glm(class~diabetes, data=df.train,family = "binomial")
inclusivefit <- glm(class~., data=df.train,family = "binomial")
bestfit <- step(inclusivefit, direction="both")

##Significance Check
anova(basefit, bestfit)

## odds ratios only
exp(coef(bestfit))

## odds ratios and 95% CI
exp(cbind(OR = coef(bestfit), confint(bestfit)))

anova(basefit, bestfit)
coefficients(bestfit) # model coefficients
confint(bestfit, level=0.95) # CIs for model parameters 
fitted(bestfit) # predicted values
anova(bestfit) # anova table 
vcov(bestfit) # covariance matrix for model parameters 
```