---
title: "CondProb"
author: "Dr. B"
date: "Monday, November 03, 2014"
output: html_document
---
Let + or - be the results of a diagnostic test, positive or negative, for a specific disease.  Let D or D' indicate if an individual has or does not have the disease.

The SENSITIVITY is the probability that the test is positive given that the individual actually has the disease, P(+|D)

The SPECIFICITY is the probability that the test is negative given that the indiviudal does not actually have the disease, P(-|D').

The positive predictive value is the probability that the individual has the disease given that the test is positive, P(D|+).

The negative predictive value is the probabilty that the subject does not have the disease given that the test is negative, P(D'|-).

The prevalence of the disease is the marginal probability of disease, P(D).

The diagnostic likelihood ratio of a positive test, DLR+, is P(+|D)/P(+|D') which is:

        sensitivity/(1-specificity)
        
The diagnostic likelihood ratio of a negative test, DLR-, is P(-|D)/P(-|D') which is:

        (1-sensitivity)/specificity
        
A test for a certain disease has a sensitivity of 99.7% and a specificity of 98.5%

Suppose that an individual from a population with a .1% prevalance of the disease receives a positive test result.  What us the positive predicitive value?

First calculate the DLR+
```{r}
##sensitivity/(1-specificity)
.997/(1-.985)
```

The results of a positive test is that the odds of disease is now 66 times the prevalance of the disease.
```{r}
((.997/(1-.985))*.001) *100
```

In other words, there is a 6.65% probability that the individual has the disease given the positive test result.



```{r}

```


```{r}

```

```{r}

```


```{r}

```


```{r}

```

```{r}

```


```{r}

```



