---
title: "CondProb"
author: "Dr. B"
date: "Monday, November 03, 2014"
output: html_document
---
Let + or - be the results of a diagnostic test, positive or negative, for a specific disease.  Let D or D' indicate if an individual has or does not have the disease.

The SENSITIVITY is the probability that the test is positive given that the individual actually has the disease, P(+|D)

The SPECIFICITY is the probability that the test is negative given that the indiviudal does not actually have the disease, P(-|D').

The positive predictive value is the probability that the individual has the disease given that the test is positive, P(D|+).

The negative predictive value is the probabilty that the subject does not have the disease given that the test is negative, P(D'|-).

The prevalence of the disease is the marginal probability of disease, P(D).

The diagnostic likelihood ratio of a positive test, DLR+, is P(+|D)/P(+|D') which is:

        sensitivity/(1-specificity)
        
The diagnostic likelihood ratio of a negative test, DLR-, is P(-|D)/P(-|D') which is:

        (1-sensitivity)/specificity
        
A test for a certain disease has a sensitivity of 99.7% and a specificity of 98.5%

Suppose that an individual from a population with a .1% prevalance of the disease receives a positive test result.  What us the positive predicitive value?

First calculate the DLR+
```{r}
##sensitivity/(1-specificity)
.997/(1-.985)
```

The results of a positive test is that the odds of disease is now 66 times the prevalance of the disease.
```{r}
((.997/(1-.985))*.001) *100
```

In other words, there is a 6.65% probability that the individual has the disease given the positive test result.



```{r}

```


```{r,warning=FALSE}
##Confidence intervals 95%
library(UsingR)
data(father.son)
x <- father.son$sheight

## Confidence Interval in inches from above data for mean in inches
(mean(x)+c(-1,1)*qnorm(.975)*sd(x)/sqrt(length(x)))

##Confidence interval for propotion assuming mean proprotion is 56% and n = 100
.56 +c(-1,1)*qnorm(.975)*sqrt(.56*.44/100)
## Or we could use the following
binom.test(56,100)$conf.int
```

```{r}
## Poisson interval
## A nuclear pump failed 5 times out of 94.32 days, give a 95% confidence interval for the failure rate per day?

x <- 5
t <- 94.32
lambda <- x/t
round(lambda + c(-1, 1) * qnorm(0.975) * sqrt(lambda/t), 3)

## Poisson exact
poisson.test(x, T = 94.32)$conf

exp(confint(glm(x ~ 1 + offset(log(t)), family = poisson(link = log))))


```


```{r,fig.height=4,fig.width=4}
## Law of large numbers in action
n <- 10000; means <- cumsum(rnorm(n)) / (1  : n)
plot(1 : n, means, type = "l", lwd = 2, 
     frame = FALSE, ylab = "cumulative means", xlab = "sample size")
abline(h = 0)
```

Sample flip of biased coin
```{r, echo = FALSE, fig.width=7.5, fig.height = 5}
par(mfrow = c(2, 3))
for (n in c(1, 10, 20)){
  temp <- matrix(sample(0 : 1, n * 10000, replace = TRUE), ncol = n)
  temp <- apply(temp, 1, mean)
  temp <- (temp - .5) * 2 * sqrt(n)
  dty <- density(temp)
  plot(dty$x, dty$y, xlab = "", ylab = "density", type = "n", xlim = c(-3, 3), ylim = c(0, .5))
  title(paste("sample mean of", n, "obs"))
  lines(seq(-3, 3, length = 100), dnorm(seq(-3, 3, length = 100)), col = grey(.8), lwd = 3)
  lines(dty$x, dty$y, lwd = 2)
}
for (n in c(1, 10, 20)){
  temp <- matrix(sample(0 : 1, n * 10000, replace = TRUE, prob = c(.9, .1)), ncol = n)
  temp <- apply(temp, 1, mean)
  temp <- (temp - .1) / sqrt(.1 * .9 / n)
  dty <- density(temp)
  plot(dty$x, dty$y, xlab = "", ylab = "density", type = "n", xlim = c(-3, 3), ylim = c(0, .5))
  title(paste("sample mean of", n, "obs"))
  lines(seq(-3, 3, length = 100), dnorm(seq(-3, 3, length = 100)), col = grey(.8), lwd = 3)
  lines(dty$x, dty$y, lwd = 2)
}
```


```{r}
```

```{r}

```


```{r}

```



