---
title: "PracMachineLearningQUiz"
author: "Dr. B"
date: "Saturday, December 13, 2014"
output: html_document
---
---
title: "Quiz 2"
author: "Liam Damewood"
date: "September 6, 2014"
output: html_document
---
### Question 1
```{r,echo=FALSE}
rm(list = ls())
```

Load the Alzheimer's disease data using the commands:

```{r}
library(AppliedPredictiveModeling)
library(caret)
data(AlzheimerDisease)
```

Which of the following commands will create training and test sets with about 50% of the observations assigned to each?

```{r}
adData = data.frame(diagnosis,predictors)
trainIndex = createDataPartition(diagnosis, p = 0.50,list=FALSE)
training = adData[trainIndex,]
testing = adData[-trainIndex,]
```

### Question 2
```{r,echo=FALSE}
rm(list = ls())
```

Load the cement data using the commands:

```{r}
library(AppliedPredictiveModeling)
data(concrete)
library(caret)
set.seed(975)
inTrain = createDataPartition(mixtures$CompressiveStrength, p = 3/4)[[1]]
training = mixtures[ inTrain,]
testing = mixtures[-inTrain,]
```

Make a plot of the outcome (CompressiveStrength) versus the index of the samples. Color by each of the variables in the data set (you may find the `cut2()` function in the Hmisc package useful for turning continuous covariates into factors). What do you notice in these plots?

```{r}
library(Hmisc)
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(Cement,g=4))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(BlastFurnaceSlag,g=4))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(FlyAsh,g=3))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(Water,g=4))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(Superplasticizer,g=4))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(CoarseAggregate,g=4))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(FineAggregate,g=4))
qplot(x = 1:length(inTrain), y = CompressiveStrength, data = training, color = cut2(Age,g=4))
```

* The data show a step like pattern that is perfectly explained by the Age variable.
* There is a step-like pattern in the plot of outcome versus index in the training set.
* There is a step-like pattern in the plot of outcome versus index in the training set that isn't explained by any of the predictor variables so there may be a variable missing.
* The data show a step like pattern that is perfectly explained by the FlyAsh variable. <--

### Question 3
```{r,echo=FALSE}
rm(list = ls())
```

Load the cement data using the commands:
```{r}
library(AppliedPredictiveModeling)
data(concrete)
library(caret)
set.seed(975)
inTrain = createDataPartition(mixtures$CompressiveStrength, p = 3/4)[[1]]
training = mixtures[ inTrain,]
testing = mixtures[-inTrain,]
```

Make a histogram and confirm the SuperPlasticizer variable is skewed. Normally you might use the log transform to try to make the data more symmetric. Why would that be a poor choice for this variable?

```{r}
qplot(Superplasticizer, data = training)
qplot(log(Superplasticizer+1), data = training)
```

* The SuperPlasticizer data include negative values so the log transform can not be performed.
* There are values of zero so when you take the log() transform those values will be -Inf. <--
* The log transform does not reduce the skewness of the non-zero values of SuperPlasticizer
* The log transform produces negative values which can not be used by some classifiers.

### Question 4
```{r,echo=FALSE}
rm(list = ls())
```

Load the Alzheimer's disease data using the commands:
```{r}
library(caret)
library(AppliedPredictiveModeling)
set.seed(3433)
data(AlzheimerDisease)
adData = data.frame(diagnosis,predictors)
inTrain = createDataPartition(adData$diagnosis, p = 3/4)[[1]]
training = adData[ inTrain,]
testing = adData[-inTrain,]
```
Find all the predictor variables in the training set that begin with IL. Perform principal components on these variables with the preProcess() function from the caret package. Calculate the number of principal components needed to capture 80% of the variance. How many are there?

```{r}
subset = training[,grep("^IL", names(training))]
preProcess(subset, thresh = 0.8, method = "pca")$numComp
```

### Question 5
```{r,echo=FALSE}
rm(list = ls())
```
Load the Alzheimer's disease data using the commands:
```{r}
library(caret)
library(AppliedPredictiveModeling)
set.seed(3433)
data(AlzheimerDisease)
adData = data.frame(diagnosis,predictors)
inTrain = createDataPartition(adData$diagnosis, p = 3/4)[[1]]
training = adData[ inTrain,]
testing = adData[-inTrain,]
```
Create a training data set consisting of only the predictors with variable names beginning with IL and the diagnosis. Build two predictive models, one using the predictors as they are and one using PCA with principal components explaining 80% of the variance in the predictors. Use method="glm" in the train function. What is the accuracy of each method in the test set? Which is more accurate?

```{r}
trainSubset = training[,grep("^IL", names(training))]
testSubset = testing[,grep("^IL", names(testing))]
pp = preProcess(trainSubset, thresh = 0.8, method = "pca")
trainTransformed <- predict(pp, trainSubset)
testTransformed <- predict(pp, testSubset)
trainSubset$diagnosis = training$diagnosis
testSubset$diagnosis = testing$diagnosis
trainTransformed$diagnosis = training$diagnosis
testTransformed$diagnosis = testing$diagnosis
glmpca = train(diagnosis ~ ., data = trainTransformed, method = "glm")
glm = train(diagnosis ~ ., data = trainSubset, method = "glm")
round(confusionMatrix(testSubset$diagnosis,predict(glm, testSubset))$overall["Accuracy"],2)
round(confusionMatrix(testTransformed$diagnosis,predict(glmpca, testTransformed))$overall["Accuracy"],2)
```


---
title: "Quiz 3"
author: "Liam Damewood"
date: "September 8, 2014"
output: html_document
---
### Question 1
Load the cell segmentation data from the AppliedPredictiveModeling package using the commands:
```{r}
library(AppliedPredictiveModeling)
data(segmentationOriginal)
library(caret)
```
1. Subset the data to a training set and testing set based on the Case variable in the data set. 
2. Set the seed to 125 and fit a CART model with the rpart method using all predictor variables and default caret settings. 
3. In the final model what would be the final model prediction for cases with the following variable values:
a. TotalIntench2 = 23,000; FiberWidthCh1 = 10; PerimStatusCh1=2 
b. TotalIntench2 = 50,000; FiberWidthCh1 = 10;VarIntenCh4 = 100 
c. TotalIntench2 = 57,000; FiberWidthCh1 = 8;VarIntenCh4 = 100 
d. FiberWidthCh1 = 8;VarIntenCh4 = 100; PerimStatusCh1=2

* a. Not possible to predict 
  b. WS 
  c. PS
  d. PS 
* a. PS 
  b. WS 
  c. PS
  d. WS 
* a. PS <--
  b. WS 
  c. PS
  d. Not possible to predict 
* a. WS 
  b. WS 
  c. PS
  d. Not possible to predict 

```{r}
training = segmentationOriginal[segmentationOriginal$Case == "Train",]
testing = segmentationOriginal[segmentationOriginal$Case == "Test",]
set.seed(125)
library(rpart)
fit = train(Class ~ ., data = training,method = "rpart")
plot(fit$finalModel, uniform = T)
text(fit$finalModel, use.n = TRUE, all = TRUE, cex = 0.8)
```

### Question 2
If K is small in a K-fold cross validation is the bias in the estimate of out-of-sample (test set) accuracy smaller or bigger? If K is small is the variance in the estimate of out-of-sample (test set) accuracy smaller or bigger. Is K large or small in leave one out cross validation?

* The bias is smaller and the variance is bigger. Under leave one out cross validation K is equal to one.
* The bias is larger and the variance is smaller. Under leave one out cross validation K is equal to the sample size. <--
* The bias is smaller and the variance is smaller. Under leave one out cross validation K is equal to the sample size.
* The bias is smaller and the variance is smaller. Under leave one out cross validation K is equal to one.

### Question 3
Load the olive oil data using the commands:
```{r}
library(pgmm)
data(olive)
olive = olive[,-1]
```
(NOTE: If you have trouble installing the __pgmm__ package, you can download the `olive` dataset here: olive_data.zip. After unzipping the archive, you can load the file using the `load()` function in R.)
These data contain information on 572 different Italian olive oils from multiple regions in Italy. Fit a classification tree where Area is the outcome variable. Then predict the value of area for the following data frame using the tree command with all defaults
```{r}
newdata = as.data.frame(t(colMeans(olive)))
```
What is the resulting prediction? Is the resulting prediction strange? Why or why not?

* 2.875. There is no reason why this result is strange.
* 4.59965. There is no reason why the result is strange.
* 0.005291005 0 0.994709 0 0 0 0 0 0. There is no reason why the result is strange.
* 2.875. It is strange because Area should be a qualitative variable - but tree is reporting the average value of Area as a numeric variable in the leaf predicted for newdata <--

```{r}
library(tree)
tr = tree(Area ~ ., data = olive)
plot(tr, uniform = T)
text(tr, use.n = TRUE, all = TRUE, cex = 0.8)
predict(tr,newdata)
```

### Question 4
Load the South Africa Heart Disease Data and create training and test sets with the following code:
```{r}
library(ElemStatLearn)
data(SAheart)
set.seed(8484)
train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
trainSA = SAheart[train,]
testSA = SAheart[-train,]
```
Then set the seed to 13234 and fit a logistic regression model (method="glm", be sure to specify family="binomial") with Coronary Heart Disease (chd) as the outcome and age at onset, current alcohol consumption, obesity levels, cumulative tabacco, type-A behavior, and low density lipoprotein cholesterol as predictors. Calculate the misclassification rate for your model using this function and a prediction on the "response" scale:
```{r}
missClass = function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}
```
What is the misclassification rate on the training set? What is the misclassification rate on the test set?

* Test Set Misclassification: 0.27 
  Training Set: 0.31
* Test Set Misclassification: 0.31 
  Training Set: 0.27
* Test Set Misclassification: 0.35 
  Training Set: 0.31
* Test Set Misclassification: 0.32 
  Training Set: 0.30

```{r}
set.seed(13234)
model = train(trainSA[,-c(10,1,4,5)], trainSA[,10], method = "glm", family = "binomial")
missClass(testSA[,10], predict(model, testSA))
missClass(trainSA[,10], predict(model, trainSA))
```

### Question 5
Load the vowel.train and vowel.test data sets:
```{r}
library(ElemStatLearn)
data(vowel.train)
data(vowel.test) 
```
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833. Fit a random forest predictor relating the factor variable y to the remaining variables. Read about variable importance in random forests here: http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm#ooberr The caret package uses by defualt the Gini importance. Calculate the variable importance using the varImp function in the caret package. What is the order of variable importance?

* The order of the variables is:
  x.2, x.1, x.5, x.8, x.6, x.4, x.3, x.9, x.7,x.10
* The order of the variables is:
  x.10, x.7, x.9, x.5, x.8, x.4, x.6, x.3, x.1,x.2
* The order of the variables is:
  x.2, x.1, x.5, x.6, x.8, x.4, x.9, x.3, x.7,x.10 <--
* The order of the variables is:
  x.1, x.2, x.3, x.8, x.6, x.4, x.5, x.9, x.7,x.10

```{r}
vowel.train$y <- as.factor(vowel.train$y)
vowel.test$y <- as.factor(vowel.test$y)
set.seed(33833)
model = train(y ~ ., data = vowel.train, method = "rf")
model$finalModel$importance[order(-model$finalModel$importance),]
varImp(model)
```


---
title: "Quiz 4"
author: "Liam Damewood"
date: "September 8, 2014"
output: html_document
---
### Question 1
Load the vowel.train and vowel.test data sets:
```{r}
library(ElemStatLearn)
data(vowel.train)
data(vowel.test) 
```
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833. Fit (1) a random forest predictor relating the factor variable y to the remaining variables and (2) a boosted predictor using the "gbm" method. Fit these both with the train() command in the caret package. 

What are the accuracies for the two approaches on the test data set? What is the accuracy among the test set samples where the two methods agree?

* RF Accuracy = 0.6061 <--
  GBM Accuracy = 0.5325 
  Agreement Accuracy = 0.6518
* RF Accuracy = 0.9881 
  GBM Accuracy = 0.8371 
  Agreement Accuracy = 0.9983
* RF Accuracy = 0.3233 
  GBM Accuracy = 0.8371 
  Agreement Accuracy = 0.9983
* RF Accuracy = 0.6061 
  GBM Accuracy = 0.6518 
  Agreement Accuracy = 0.5325

```{r}
library(caret)
vowel.train$y <- as.factor(vowel.train$y)
vowel.test$y <- as.factor(vowel.test$y)
set.seed(33833)
modelRF <- train(y ~ ., data = vowel.train, method = "rf")
modelGBM <- train(y ~ ., data = vowel.train, method = "gbm", verbose = F)
sum((predict(modelRF, vowel.test) == vowel.test$y)*1) / length(vowel.test$y)
sum((predict(modelGBM, vowel.test) == vowel.test$y)*1) / length(vowel.test$y)
sum((predict(modelRF, vowel.test) == predict(modelGBM, vowel.test))*1) / length(vowel.test$y)
qplot(predict(modelGBM, vowel.test),y,data=vowel.test) + geom_jitter()
```

### Question 2
Load the Alzheimer's data using the following commands
```{r}
library(caret)
library(gbm)
set.seed(3433)
library(AppliedPredictiveModeling)
data(AlzheimerDisease)
adData = data.frame(diagnosis,predictors)
inTrain = createDataPartition(adData$diagnosis, p = 0.75)[[1]]
training = adData[ inTrain,]
testing = adData[-inTrain,]
```
Set the seed to 62433 and predict diagnosis with all the other variables using a random forest ("rf"), boosted trees ("gbm") and linear discriminant analysis ("lda") model. Stack the predictions together using random forests ("rf"). What is the resulting accuracy on the test set? Is it better or worse than each of the individual predictions?

* Stacked Accuracy: 0.76 is better than lda but not random forests or boosting.
* Stacked Accuracy: 0.93 is better than all three other methods
* Stacked Accuracy: 0.79 is better than random forests and lda and the same as boosting.
* Stacked Accuracy: 0.79 is better than all three other methods <--

```{r}
set.seed(62433)
modelRF <- train(diagnosis ~ ., data = training, method = "rf")
modelGBM <- train(diagnosis ~ ., data = training, method = "gbm", verbose = F)
modelLDA <- train(diagnosis ~ ., data = training, method = "lda")
A = max(modelRF$results$Accuracy)
B = max(modelGBM$results$Accuracy)
C = max(modelLDA$results$Accuracy)
A*B*C+(1-A)*B*C+A*(1-B)*C+A*B*(1-C)
```

### Question 3
Load the concrete data with the commands:
```{r}
set.seed(3523)
library(AppliedPredictiveModeling)
data(concrete)
inTrain = createDataPartition(concrete$CompressiveStrength, p = 3/4)[[1]]
training = concrete[ inTrain,]
testing = concrete[-inTrain,]
```
Set the seed to 233 and fit a lasso model to predict Compressive Strength. Which variable is the last coefficient to be set to zero as the penalty increases? (Hint: it may be useful to look up ?plot.enet).

* Cement <--
* Age
* Water
* FineAggregate

```{r}
set.seed(233)
tg = data.frame(fraction = (1:99)/100)
model = train(CompressiveStrength ~ ., data = concrete, method = "lasso", tuneGrid = tg)
plot(model$finalModel, xvar = "penalty", use.color = T)
```

### Question 4
Load the data on the number of visitors to the instructors blog from here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/gaData.csv
```{r}
download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/gaData.csv", destfile = "~/Desktop/gaData.csv")
```
Using the commands:

```{r}
library(lubridate)  # For year() function below
dat = read.csv("~/Desktop/gaData.csv")
training = dat[year(dat$date) < 2012,]
testing = dat[(year(dat$date)) > 2011,]
tstrain = ts(training$visitsTumblr)
```
Fit a model using the bats() function in the forecast package to the training time series. Then forecast this model for the remaining time points. For how many of the testing points is the true value within the 95% prediction interval bounds?

* 96% <--
* 93%
* 92%
* 98%

```{r}
library(forecast)
model = bats(tstrain)
fore = forecast(model, 235)
lo = fore$lower[,2]
hi = fore$upper[,2]
round(sum(testing[,3] <= hi & testing[,3] >= lo) / length(lo),2)*100
```

### Question 5
Load the concrete data with the commands:
```{r}
set.seed(3523)
library(AppliedPredictiveModeling)
data(concrete)
inTrain = createDataPartition(concrete$CompressiveStrength, p = 3/4)[[1]]
training = concrete[ inTrain,]
testing = concrete[-inTrain,]
```
Set the seed to 325 and fit a support vector machine using the e1071 package to predict Compressive Strength using the default settings. Predict on the testing set. What is the RMSE?

* 6.93
* 6.72
* 35.59 <--
* 107.44

```{r}
set.seed(325)
library(e1071)
model = svm(CompressiveStrength ~ ., data = concrete)
sqrt(sum((predict(model, testing) - testing$CompressiveStrength)^2))
```