---
title: "Graphs"
author: "Dr. B"
date: "Thursday, September 25, 2014"
output: html_document
---

```{r}
# Define the cars vector with 5 values and a trucks vector with 5 values
cars <- c(1, 3, 6, 4, 9)
trucks <- c(2, 5, 4, 5, 12)
suvs <- c(4,4,6,6,16)
```

Line Charts
-----------
A very simple graph using the values in the car vector: 
```{r}
# Graph the cars vector with all defaults
plot(cars)
```

Add a title, a line to connect the points, and some color: 
```{r}
# Graph cars using blue points overlayed by a line 
plot(cars, type="o", col="blue")

# Create a title with a red, bold/italic font
title(main="Autos", col.main="red", font.main=4)
```


Add a red line for trucks and specify the y-axis range directly so it will be large enough to fit the truck data: 
```{r}
# Graph cars using a y axis that ranges from 0 to 12
plot(cars, type="o", col="blue", ylim=c(0,12))

# Graph trucks with red dashed line and square points
lines(trucks, type="o", pch=22, lty=2, col="red")

# Create a title with a red, bold/italic font
title(main="Autos & Trucks", col.main="red", font.main=4)

```

Change the axes labels to match our data and add a legend. We'll also compute the y-axis values using the max function so any changes to our data will be automatically reflected in our graph. 
```{r}
# Calculate range from 0 to max value of cars and trucks
g_range <- range(0, cars, trucks)

# Graph autos using y axis that ranges from 0 to max 
# value in cars or trucks vector.  Turn off axes and 
# annotations (axis labels) so we can specify them ourself
plot(cars, type="o", col="blue", ylim=g_range, 
   axes=FALSE, ann=FALSE)

# Make x axis using Mon-Fri labels
axis(1, at=1:5, lab=c("Mon","Tue","Wed","Thu","Fri"))

# Make y axis with horizontal labels that display ticks at 
# every 4 marks. 4*0:g_range[2] is equivalent to c(0,4,8,12).
axis(2, las=1, at=4*0:g_range[2])

# Create box around plot
box()

# Graph trucks with red dashed line and square points
lines(trucks, type="o", pch=22, lty=2, col="red")

# Create a title with a red, bold/italic font
title(main="Autos", col.main="red", font.main=4)

# Label the x and y axes with dark green text
title(xlab="Days", col.lab=rgb(0,0.5,0))
title(ylab="Total", col.lab=rgb(0,0.5,0))

# Create a legend at (1, g_range[2]) that is slightly smaller 
# (cex) and uses the same line colors and points used by 
# the actual plots 
legend(1, g_range[2], c("cars","trucks"), cex=0.8, 
   col=c("blue","red"), pch=21:22, lty=1:2);
```

Write a graph to a png file
```{r}
# Compute the largest y value used in the data (or we could
# just use range again)
autos <- c(cars, trucks, suvs)
max_y <- max(autos)

# Define colors to be used for cars, trucks, suvs
plot_colors <- c("blue","red","forestgreen")

# Start PNG device driver to save output to figure.png
png(filename="figure.png", height=295, width=300, bg="white")

# Graph autos using y axis that ranges from 0 to max_y.
# Turn off axes and annotations (axis labels) so we can 
# specify them ourself
plot(cars, type="o", col=plot_colors[1], ylim=c(0,max_y), axes=FALSE, ann=FALSE)

# Make x axis using Mon-Fri labels
axis(1, at=1:5, lab=c("Mon", "Tue", "Wed", "Thu", "Fri"))

# Make y axis with horizontal labels that display ticks at 
# every 4 marks. 4*0:max_y is equivalent to c(0,4,8,12).
axis(2, las=1, at=4*0:max_y)

# Create box around plot
box()

# Graph trucks with red dashed line and square points
lines(trucks, type="o", pch=22, lty=2, col=plot_colors[2])

# Graph suvs with green dotted line and diamond points
lines(suvs, type="o", pch=23, lty=3, col=plot_colors[3])

# Create a title with a red, bold/italic font
title(main="Autos", col.main="red", font.main=4)

# Label the x and y axes with dark green text
title(xlab= "Days", col.lab=rgb(0,0.5,0))
title(ylab= "Total", col.lab=rgb(0,0.5,0))

# Create a legend at (1, max_y) that is slightly smaller 
# (cex) and uses the same line colors and points used by 
# the actual plots
legend(1, max_y, c("cars","trucks","suvs"), cex=0.8, col=plot_colors, 
   pch=21:23, lty=1:3);
   
# Turn off device driver (to flush output to png)
dev.off()

```

Write a graph to a pdf file
```{r}
   		
# Define colors to be used for cars, trucks, suvs
plot_colors <- c(rgb(r=0.0,g=0.0,b=0.9), "red", "forestgreen")

# Start PDF device driver to save output to figure.pdf
pdf(file="figure.pdf", height=3.5, width=5)
			
# Trim off excess margin space (bottom, left, top, right)
par(mar=c(4.2, 3.8, 0.2, 0.2))
			
# Graph autos using a y axis that uses the full range of value
# in autos_data. Label axes with smaller font and use larger 
# line widths.
plot(cars, type="l", col=plot_colors[1], ylim=c(0,max_y), axes=F, ann=T, xlab="Days",
   ylab="Total", cex.lab=0.8, lwd=2)

# Make x axis tick marks without labels
axis(1, lab=F)

# Plot x axis labels at default tick marks with labels at 
# 45 degree angle
text(axTicks(1), par("usr")[3] - 2, srt=45, adj=1,
          labels=c("Mon", "Tue", "Wed", "Thu", "Fri"),
          xpd=T, cex=0.8)

# Plot y axis with smaller horizontal labels 
axis(2, las=1, cex.axis=0.8)

# Create box around plot
box()

# Graph trucks with thicker red dashed line
lines(trucks, type="l", lty=2, lwd=2, col=plot_colors[2])

# Graph suvs with thicker green dotted line
lines(suvs, type="l", lty=3, lwd=2,col=plot_colors[3])

# Create a legend in the top-left corner that is slightly  
# smaller and has no border
legend("topleft", c("cars","trucks","suvs"), cex=0.8, col=plot_colors, 
   lty=1:3, lwd=2, bty="n");
  
# Turn off device driver (to flush output to PDF)
dev.off()

# Restore default margins
par(mar=c(5, 4, 4, 2) + 0.1)

```
Bar Charts
----------
Let's start with a simple bar chart graphing the cars vector: 
```{r}
# Graph the cars vector with all defaults
barplot(cars)
```

Add labels, blue borders around the bars, and density lines: 
```{r}
# Graph cars with specified labels for axes.  Use blue 
# borders and diagnal lines in bars.
barplot(cars, main="Cars", xlab="Days",  
   ylab="Total", names.arg=c("Mon","Tue","Wed","Thu","Fri"), 
   border="blue", density=c(10,20,30,40,50))
```

Graph the total number of autos per day using some color and show a legend: 
```{r}
# Graph autos with adjacent bars using rainbow colors
barplot(c(cars,trucks,suvs), main="Autos", ylab= "Total",beside=TRUE, col=rainbow(5))

# Place the legend at the top-left corner with no frame  
# using rainbow colors
legend("topleft", c("Mon","Tue","Wed","Thu","Fri"), cex=0.6, bty="n", fill=rainbow(5));

```

Graph the total number of autos per day using a stacked bar chart and place the legend outside of the plot area: 
```{r}
# Expand right side of clipping rect to make room for the legend
par(xpd=T, mar=par()$mar+c(0,0,0,4))

# Graph autos (transposing the matrix) using heat colors,  
# put 10% of the space between each bar, and make labels  
# smaller with horizontal y-axis labels
barplot(t(cars), main="Autos", ylab="Total", 
   col=heat.colors(3), space=0.1, cex.axis=0.8, las=1,
   names.arg=c("Mon","Tue","Wed","Thu","Fri"), cex=0.8) 
   
# Place the legend at (6,30) using heat colors
legend(6, 30, c("cars","trucks","suvs"), cex=0.8, fill=heat.colors(3));
   
# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)
```

Histograms
---------
Basic Histogram
```{r}
# Create a histogram for suvs
hist(suvs)
```

plot a histogram of the combined car, truck, and suv data in color. 
```{r}
# Concatenate the three vectors
autos <- c(cars, trucks, suvs)

# Create a histogram for autos in light blue with the y axis
# ranging from 0-10
hist(autos, col="lightblue", ylim=c(0,10))
```

Change the breaks so none of the values are grouped together and flip the y-axis labels horizontally. 
```{r}
# Concatenate the three vectors
autos <- c(cars, trucks, suvs)

# Compute the largest y value used in the autos
max_num <- max(autos)

# Create a histogram for autos with fire colors, set breaks
# so each number is in its own group, make x axis range from
# 0-max_num, disable right-closing of cell intervals, set
# heading, and make y-axis labels horizontal
hist(autos, col=heat.colors(max_num), breaks=max_num, 
   xlim=c(0,max_num), right=F, main="Autos Histogram", las=1)

```

Create uneven breaks and graph the probability density.
```{r}
# Concatenate the three vectors
autos <- c(cars, trucks, suvs)

# Compute the largest y value used in the autos
max_num <- max(autos)

# Create uneven breaks
brk <- c(0,3,4,5,6,10,16)

# Create a histogram for autos with fire colors, set uneven
# breaks, make x axis range from 0-max_num, disable right-
# closing of cell intervals, set heading, make y-axis labels 
# horizontal, make axis labels smaller, make areas of each
# column proportional to the count
hist(autos, col=heat.colors(length(brk)), breaks=brk, 
   xlim=c(0,max_num), right=F, main="Probability Density", 
   las=1, cex.axis=0.8, freq=F)
```

Pie Chart
---------
```{r}
# Create a pie chart for cars
pie(cars)
```

Add a heading, change the colors, and define our own labels: 
```{r}
# Create a pie chart with defined heading and
# custom colors and labels
pie(cars, main="Cars", col=rainbow(length(cars)),
   labels=c("Mon","Tue","Wed","Thu","Fri"))

```

Change the colors, label using percentages, and create a legend: 
```{r}
# Define some colors ideal for black & white print
colors <- c("white","grey70","grey90","grey50","black")

# Calculate the percentage for each day, rounded to one 
# decimal place
car_labels <- round(cars/sum(cars) * 100, 1)

# Concatenate a '%' char after each value
car_labels <- paste(car_labels, "%", sep="")

# Create a pie chart with defined heading and custom colors
# and labels
pie(cars, main="Cars", col=colors, labels=car_labels,
   cex=0.8)

# Create a legend at the right   
legend(1.5, 0.5, c("Mon","Tue","Wed","Thu","Fri"), cex=0.8, 
   fill=colors)
```