---
title: "MarketBasket"
author: "Bryan Aguiar"
date: "Friday, May 01, 2015"
output: html_document
---


```{r,warning=FALSE,message=FALSE}

# Clear the environment
rm(list=ls())

# Turn off scientific notations for numbers
options(scipen = 999)  

# Set locale
Sys.setlocale("LC_ALL", "English") 

# Set seed for reproducibility
set.seed(2345)

# Load the libraries
library(arules)
library(arulesViz)
library(datasets)

# Load the Data
data(Groceries)
```

With this we can understand three important ratios; the support, confidence and lift. We describe the significance of these in the following bullet points, but if you are interested in a formal mathematical definition you can find it on wikipedia.

    Support: The fraction of which our item set occurs in our dataset.
    
    Confidence: probability that a rule is correct for a new transaction with items on the left.
    
    Lift: The ratio by which by the confidence of a rule exceeds the expected confidence. Note: if the lift is 1 it indicates that the items on the left and right are independent.

##Apriori Recommendation with R

View the data before making the rules.
```{r, echo=FALSE}
# Create an item frequency plot for the top 20 items
itemFrequencyPlot(Groceries,topN=20,type="absolute")
```

### Get the rules

You will always have to pass the minimum required support and confidence.
  
      We set the minimum support to 0.001
    
      We set the minimum confidence of 0.8
    
      We then show the top 5 rules

```{r}
rules <- apriori(Groceries, parameter = list(supp = 0.001, conf = 0.8))
 
# Show the top 5 rules, but only 2 digits
options(digits=2)
inspect(rules[1:5])
```

This reads easily, for example: if someone buys yogurt and cereals, they are 81% likely to buy whole milk too.  We can get summary info. about the rules that give us some interesting information such as:
```{r}
summary(rules)
```

      The number of rules generated: 410
    
      The distribution of rules by length: Most rules are 4 items long
    
      The summary of quality measures: interesting to see ranges of support, lift, and confidence.
    
      The information on the data mined: total data mined, and minimum parameters.

The first issue we see here is that the rules are not sorted. Often we will want the most relevant rules first. Lets say we wanted to have the most likely rules. We can easily sort by confidence:
```{r}
rules<-sort(rules, by="confidence", decreasing=TRUE)
inspect(rules[1:5])
```

Rule 4 is perhaps excessively long. Lets say you wanted more concise rules. That is also easy to do by adding a "maxlen" parameter to your apriori function:
```{r,warning=FALSE,message=FALSE}
rules <- apriori(Groceries, parameter = list(supp = 0.001, conf = 0.8,maxlen=3))
rules<-sort(rules, by="confidence", decreasing=TRUE)
inspect(rules[1:5])
```

Sometimes, rules will repeat. Redundancy indicates that one item might be a given. As an analyst you can elect to drop the item from the dataset. Alternatively, you can remove redundant rules generated:
```{r,warning=FALSE,message=FALSE}
subset.matrix <- is.subset(rules, rules)
subset.matrix[lower.tri(subset.matrix, diag=T)] <- NA
redundant <- colSums(subset.matrix, na.rm=T) >= 1
rules.pruned <- rules[!redundant]
rules<-rules.pruned
rules<-sort(rules, by="confidence", decreasing=TRUE)
inspect(rules[1:5])
```

###Targeting Items
Now that we know how to generate rules, limit the output, lets say we wanted to target items to generate rules. There are two types of targets we might be interested in that are illustrated with an example of "whole milk":

      What are customers likely to buy before buying whole milk
      
      What are customers likely to buy if they purchase whole milk?

This essentially means we want to set either the Left Hand Side and Right Hand Side.  Answering the first question we adjust our apriori() function as follows:
```{r,warning=FALSE,warning=FALSE}
rules<-apriori(data=Groceries, parameter=list(supp=0.001,conf = 0.08), appearance = list(default="lhs",rhs="whole milk"),             control = list(verbose=F))
rules<-sort(rules, decreasing=TRUE,by="confidence")
inspect(rules[1:5])
```

Likewise, we can set the left hand side to be "whole milk" and find its antecedents:
      
      We set the confidence to 0.15 since we get no rules with 0.8
      
      We set a minimum length of 2 to avoid empty left hand side items
```{r,warning=FALSE,message=FALSE}
rules<-apriori(data=Groceries, parameter=list(supp=0.001,conf = 0.15,minlen=2),appearance = list(default="rhs",lhs="whole milk"),
               control = list(verbose=F))
rules<-sort(rules, decreasing=TRUE,by="confidence")
inspect(rules[1:5])
```

###Visualization
The last step is visualization. Lets say you wanted to map out the rules in a graph. 
```{r,warning=FALSE,message=FALSE,echo=FALSE}
plot(rules,method="graph",interactive=TRUE,shading=TRUE)
```
