---
title: "RandomForest"
author: "Dr.B"
date: "Sunday, May 24, 2015"
output: html_document
---
```{r,warning=FALSE,message=FALSE}
# Load functions
source('functions.R')

# Load the libraries
require(caret)
require(ggplot2)
require(randomForest)
```

Read in the Training and Test Set:
```{r,warning=FALSE,message=FALSE}
# Load the data
training_URL<-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_URL<-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training<-read.csv(training_URL,na.strings=c("NA",""))
testing<-read.csv(test_URL,na.strings=c("NA",""))
```

Get rid of the columns that is simply an index, timestamp or username.
```{r,warning=FALSE,message=FALSE}
training<-training[,7:160]
testing<-testing[,7:160]
```

Remove the columns that are mostly NAs. They could be useful in the model, but it is easier to cut the data.frame down and see if it gives good results
```{r,warning=FALSE,message=FALSE}
mostly_data<-apply(!is.na(training),2,sum)>19621
training<-training[,mostly_data]
testing<-testing[,mostly_data]
```

Set training control
```{r,warning=FALSE,message=FALSE}
fitControl <- trainControl(method = "repeatedcv",number =2,repeats = 1)
```

Partition the training set into a smaller set to speed up the running of the model
```{r,warning=FALSE,message=FALSE}
InTrain<-createDataPartition(y=training$classe,p=0.1,list=FALSE)
training1<-training[InTrain,]
```

Train the model
```{r,warning=FALSE,message=FALSE}
rf_model<-train(classe~.,data=training1,method="rf",trControl=fitControl,prox=TRUE,allowParallel=TRUE)
print(rf_model)
```

```{r,warning=FALSE,message=FALSE}
print(rf_model$finalModel)
pred<-predict(rf_model$finalModel,testing)
summary(pred)

plot(rf_model)
plot(rf_model$finalModel)

plot(varImp(rf_model),top=10)
````


