---
title: "Statistical Inference Quizzes"
author: "Dr. B"
date: "Monday, November 24, 2014"
output: html_document
---
##Quiz 1

###1
Consider influenza epidemics for two parent heterosexual families. Suppose that the probability is 17% that at least one of the parents has contracted the disease. The probability that the father has contracted influenza is 12% while the probability that both the mother and father have contracted the disease is 6%. What is the probability that the mother has contracted influenza?(0.11)

```{r}
paub <- 0.17
pa <- 0.12
panb <- 0.06
pb <- paub + panb - pa
pb 
```

###2
A random variable, X is uniform, a box from 0 to 1 of height 1. (So that its density is f(x)=1 for 0???x???1.) What is its 75th percentile? (0.75)

```{r}
pbeta(0.75, 1, 1) 
```

###3
You are playing a game with a friend where you flip a coin and if it comes up heads you give her X dollars and if it comes up tails she gives you Y dollars. The probability that the coin is heads is p (some number between 0 and 1.) What has to be true about X and Y to make so that both of your expected total earnings is 0. The game would then be called "fair". (p/(1-p) = Y/X)
```{r}
##p*(-X) + (1-p)*Y=0
##p*X = (1-p)*Y
##p/(1-p) = Y/X
```

###4
A density that looks like a normal density (but may or may not be exactly normal) is exactly symmetric about zero. (Symmetric means if you flip it around zero it looks the same.) What is its median? (1)

```{r}
ex <- 1/2*(1-1)
var <- (1^2+1^2)/2
var 
```

###5
```{r}

```


```{r}

```


```{r}

```


```{r}

```
