---
title: "T-Test"
author: "Dr. B"
date: "Thursday, November 13, 2014"
output: html_document
---

A few different T Tests
```{r,warning=FALSE}
##Use my standard openning including call function
if (Sys.info()["sysname"]=="Linux"){
  source('/home/bryan/git/MyWork/StdOpen.R')     
}else{
  source('C:/Users/bryan_000/Documents/GitHub/MyWork/StdOpen.R')   
}

call("reshape2")
data(ToothGrowth)

##Make dose a factor
ToothGrowth$dose = factor(ToothGrowth$dose, levels=c(0.5,1.0,2.0),labels=c("low","med","high"))

##SPli the datset into testing and training
split <-dfsplit(ToothGrowth)
df.test <-split$testset
df.train <-split$trainset
```

The ToothGrowth dataset explains the relation between the length of teeth in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1 and 2 mg) with each of two supplement types (orange juice and ascorbic acid). 

First, we make a boxplot
```{r,echo=FALSE}
with(ToothGrowth, {
    boxplot(len ~ dose, boxwex = 0.25, at = 1:3 - 0.2,
            subset = (supp == "VC"), col = "yellow",
            main = "Guinea Pigs' Tooth Growth",
            xlab = "Vitamin C dose mg",
            ylab = "tooth length", ylim = c(0, 35))
    boxplot(len ~ dose, add = TRUE, boxwex = 0.25, at = 1:3 + 0.2,
            subset = supp == "OJ", col = "orange")
    legend(2, 9, c("Ascorbic acid", "Orange juice"),
           fill = c("yellow", "orange"))
})
```

Next, we make a table
```{r}
table(ToothGrowth$dose,ToothGrowth$supp)
```

##Summaries of the Data
Next, we will look at the mean, variance, and standard deviation of tooth length, grouped by supplement type and dose.
```{r}
###Basic Summary of the Data 
summaryBy(len~supp+dose, data=ToothGrowth, FUN=c(mean,var,sd,length))

##Mean
with(ToothGrowth, tapply(len, list(dose,supp), mean))

##Variance
with(ToothGrowth, tapply(len, list(dose,supp), var))

##SD
with(ToothGrowth, tapply(len, list(dose,supp), sd))

##Counts
with(ToothGrowth, tapply(len, list(dose,supp), length))
```

##Hypothesis Tests and Confidence Intervals
####Paired By Supplement
The hypothesis test was that there was no difference in tooth length between orange juice and ascorbic acid.  The alternative hypothesis was that orange juice increased tooth length more than ascorbic acid did.
```{r}
results <-t.test(ToothGrowth$len[ToothGrowth$supp=="OJ"],ToothGrowth$len[ToothGrowth$supp=="VC"],paired=TRUE, alternative = "greater")
results
```

####Paired By Dose (High versus Medium)
The null hypothesis was that there was no difference in tooth length between 2 mg and 1 mg dose levels.  The alternative hypothesis was that the 2 mg dose increased tooth length more than 1 mg dose did.
```{r}
results <-t.test(ToothGrowth$len[ToothGrowth$dose=="high"],ToothGrowth$len[ToothGrowth$dose=="med"],paired=TRUE, alternative = "greater")
results
```

####Paired By Dose (Medium versus Low)
The null hypothesis was that there was no difference in tooth length between the 1 mg and .5 mg dose levels.  The alternative hypothesis was that the 1 mg dose increased tooth length more than .5 mg dose did.
```{r}
results <-t.test(ToothGrowth$len[ToothGrowth$dose=="med"],ToothGrowth$len[ToothGrowth$dose=="low"],paired=TRUE, alternative = "greater")
results
```

##Group Dose by Supplement Type
We extract the data by supplement type, creating two new data frames, so we can run various T-tests. 
```{r}
## Pull out dosage by supplement
vcdf <-subset (ToothGrowth,supp=="VC")
ojdf <-subset (ToothGrowth,supp=="OJ")
```

###Paired High Dose
The null hypothesis was that there was no difference in tooth length between the dose levels by supplement type.  The alternative hypothesis was that the at the high dose, orange juice supplement increased tooth length more than ascorbic acid did.
```{r}
## Check High dosages
results <-t.test(ojdf$len[ojdf$dose=="high"],vcdf$len[vcdf$dose=="high"],paired=TRUE, alternative = "greater")
results
```

###Paired Medium Dose
The null hypothesis was that there was no difference in tooth length between the dose levels by supplement type.  The alternative hypothesis was that the at the medium dose, orange juice supplement increased tooth length more than ascorbic acid did.
```{r}
## Check Medium dosages
results <-t.test(ojdf$len[ojdf$dose=="med"],vcdf$len[vcdf$dose=="med"],paired=TRUE, alternative = "greater")
results
```

###Paired Low Dose
The null hypothesis was that there was no difference in tooth length between the dose levels by supplement type.  The alternative hypothesis was that the at the low dose, orange juice supplement increased tooth length more than ascorbic acid did.
```{r}
## Check Low dosages
results <-t.test(ojdf$len[ojdf$dose=="low"],vcdf$len[vcdf$dose=="low"],paired=TRUE, alternative = "greater")
results
```

##Independent T-Tests
###Independent (assume equal var) By Dose (High versus Medium)
The null hypothesis was that there was no difference in tooth length between 2 mg and 1 mg dose levels.  The alternative hypothesis was that the 2 mg dose increased tooth length more than 1 mg dose did.
```{r}
results <-t.test(ToothGrowth$len[ToothGrowth$dose=="high"],ToothGrowth$len[ToothGrowth$dose=="med"],var.equal=TRUE, alternative = "greater")
results
```

###Independent (assume equal var) By Dose (Medium versus Low)
The null hypothesis was that there was no difference in tooth length between 2 mg and 1 mg dose levels.  The alternative hypothesis was that the 1 mg dose increased tooth length more than .5 mg dose did.
```{r}
results <-t.test(ToothGrowth$len[ToothGrowth$dose=="med"],ToothGrowth$len[ToothGrowth$dose=="low"],var.equal=TRUE, alternative = "greater")
results
```

##Analysis of Variation
###One-Way Anova by Dose
```{r}
output <- aov(len ~ dose, data=ToothGrowth)
model.tables(output,"means",se=T)
summary(output)
plot(output)
```

###Post Hoc Tests
####Pairwise T Test
The function pairwise.t.test computes the pair-wise comparisons between group means with corrections for multiple testing. The general form is pairwise.t.test(response, factor, p.adjust = method, alternative = c("two.sided","less", "greater"))
```{r}
pairwise.t.test(ToothGrowth$len, ToothGrowth$dose,  p.adjust="bonferroni")
```

###Anova by Supplement & Dose
####ANOVA
```{r}
output <- aov(len ~ supp + dose, data=ToothGrowth)
model.tables(output,"means",se=T)
summary(output)
plot(output)
TukeyHSD(output)
```

####ANOVA with Interaction
```{r}
output <- aov(len ~ supp*dose,data=ToothGrowth )
model.tables(output,"means", se=T)
summary(output)
plot(output)
TukeyHSD(output)

##Interaction Plt
with(ToothGrowth, interaction.plot(x.factor=dose, trace.factor=supp,response=len, fun=mean, type="b", legend=T, ylab="Tooth Length", main="Interaction Plot", pch=c(1,19)))

##Interaction Plot and 
with(ToothGrowth,
interaction.plot(dose, supp, len, fun=mean,
xlab="Dose", ylab="Average Length",
lty=1, lwd=2, col=c("red", "blue"), legend=FALSE))
legend("bottomright", c("Orange Juice", "Ascorbic Acid"),
col=c("red", "blue"), lty=1, bty="n", title="Supplement")

```

##Linear Model
```{r}
fit <- aov(len~dose+supp, data=ToothGrowth)
summary(fit)
```


```{r}
##Make a 
x <- ToothGrowth$len[ToothGrowth$dose=="high"]-ToothGrowth$len[ToothGrowth$dose=="med"]
summary(x)
results <-t.test(x,  alternative = "greater")
results

x <- ToothGrowth$len[ToothGrowth$dose=="med"]-ToothGrowth$len[ToothGrowth$dose=="low"]
summary(x)
results <-t.test(x,  alternative = "greater")
results

x <- ToothGrowth$len[ToothGrowth$dose=="high"]-ToothGrowth$len[ToothGrowth$dose=="low"]
summary(x)
results <-t.test(x,  alternative = "greater")
results

x <- ToothGrowth$len[ToothGrowth$supp=="OJ"]-ToothGrowth$len[ToothGrowth$supp=="VC"]
summary(x)
results <-t.test(x,  alternative = "greater")
results
```

### One-Sample T-Tests
```{r}
# Add the ID column to the data frame
dataIDAdded <- data.frame(ToothGrowth,PigID=rep(c(1:10),6))
tgdata <- dcast(dataIDAdded,dose+PigID~supp,value.var="len")

# Adjust the column names of the data frame
names(tgdata)[-(1: 2)] <- paste("supp_", names(tgdata)[-(1: 2)], sep = "")

# Add a difference value column to the data frame
tgdata <- mutate(tgdata,diffval=supp_OJ-supp_VC)

##
tgdata05 <- subset(tgdata,dose=="low")
t.test(tgdata05$diffval)

tgdata10 <- subset(tgdata,dose=="med")
t.test(tgdata10$diffval)

tgdata20 <- subset(tgdata,dose=="high")
t.test(tgdata20$diffval)
```
